---
title: "Assignment 2 - Language Development in ASD - Part 1 - Explaining development"
author: "[YOUR NAME]"
date: "[DATE]"
output: html_document
---
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(include = FALSE)
```

# Assignment 2

In this assignment you will have to discuss a few important questions (given the data you have). More details below. The assignment submitted to the teachers consists of:
- a report answering and discussing the questions (so we can assess your conceptual understanding and ability to explain and critically reflect)
- a link to a git repository with all the code (so we can assess your code)

Part 1 - Basic description of language development
- Describe your sample (n, age, gender, clinical and cognitive features of the two groups) and critically assess whether the groups (ASD and TD) are balanced
- Describe linguistic development (in terms of MLU over time) in TD and ASD children (as a function of group). 
- Describe how parental use of language (in terms of MLU) changes over time. What do you think is going on?
- Include individual differences in your model of language development (in children). Identify the best model.

Part 2 - Model comparison
- Discuss the differences in performance of your model in training and testing data
- Which individual differences should be included in a model that maximizes your ability to explain/predict new data?
- Predict a new kid's performance (Bernie) and discuss it against expected performance of the two groups

Part 3 - Simulations to plan a new study
- Report and discuss a power analyses identifying how many new kids you would need to replicate the results

The following involves only Part 1.

## Learning objectives

- Summarize and report data and models
- Critically apply mixed effects (or multilevel) models
- Explore the issues involved in feature selection


# Quick recap
Autism Spectrum Disorder is often related to language impairment. However, this phenomenon has not been empirically traced in detail:
i) relying on actual naturalistic language production,  ii) over extended periods of time.

We therefore videotaped circa 30 kids with ASD and circa 30 comparison kids (matched by linguistic performance at visit 1) for ca. 30 minutes of naturalistic interactions with a parent. We repeated the data collection 6 times per kid, with 4 months between each visit. We transcribed the data and counted: 
i) the amount of words that each kid uses in each video. Same for the parent.
ii) the amount of unique words that each kid uses in each video. Same for the parent.
iii) the amount of morphemes per utterance (Mean Length of Utterance) displayed by each child in each video. Same for the parent. 

This data is in the file you prepared in the previous class. 

NB. A few children have been excluded from your datasets. We will be using them next week to evaluate how good your models are in assessing the linguistic development in new participants.

This RMarkdown file includes 
1) questions (see above). Questions have to be answered/discussed in a separate document that you have to directly send to the teachers.
2) A break down of the questions into a guided template full of hints for writing the code to solve the exercises. Fill in the code and the paragraphs as required. Then report your results in the doc for the teachers.

REMEMBER that you will have to have a github repository for the code and send the answers to Kenneth and Riccardo without code (but a link to your github/gitlab repository). This way we can check your code, but you are also forced to figure out how to report your analyses :-)

Before we get going, here is a reminder of the issues you will have to discuss in your report:

1- Describe your sample (n, age, gender, clinical and cognitive features of the two groups) and critically assess whether the groups (ASD and TD) are balanced
2- Describe linguistic development (in terms of MLU over time) in TD and ASD children (as a function of group). 
3- Describe how parental use of language (in terms of MLU) changes over time. What do you think is going on?
4- Include individual differences in your model of language development (in children). Identify the best model.

# Let's go

### Loading the relevant libraries

Load necessary libraries : what will you need?
- e.g. something to deal with the data
- e.g. mixed effects models
- e.g. something to plot with

```{r Load Libraries, include = FALSE}
library(pacman)

p_load(tidyverse, lmerTest, ggplot2, MuMINn)

install.packages("MuMIn")

library(MuMINn)
```

### Define your working directory and load the data
If you created a project for this class and opened this Rmd file from within that project, your working directory is your project directory.

If you opened this Rmd file outside of a project, you will need some code to find the data:
- Create a new variable called locpath (localpath)
- Set it to be equal to your working directory
- Move to that directory (setwd(locpath))
- Load the data you saved last time (use read_csv(fileName))

```{r Load Data, include = FALSE}
# the directory is already placed in a project, where the data is placed

# Loading the data
train <- read.csv('train.csv', header = T, sep =',')

```

### Characterize the participants (Exercise 1)

Identify relevant variables: participants demographic characteristics, diagnosis, ADOS, Verbal IQ, Non Verbal IQ, Socialization, Visit, Number of words used, Number of unique words used, mean length of utterance in both child and parents.

Make sure the variables are in the right format.

Describe the characteristics of the two groups of participants and whether the two groups are well matched.

```{r descriptive stats, include = FALSE}
#making subject charectors in stead of a numeric value
train$SUBJ <- as.character(train$SUBJ)

summary(train)


# Creating a summary of our two different groups, ASD and TD
# This is to be discriped in the report 
# Age is in months 

train %>% subset(VISIT == 1) %>%
  group_by(Diagnosis) %>%
  dplyr::summarise('Number' = n(), 
                   'Female' = sum(Gender == 'F'),
                   'Male' = sum(Gender == 'M'),
                   'Age' = mean(Age, na.rm = T),
                   'ADOS' = mean(ADOS),
                   'Verbal IQ' = mean(ExpressiveLangRaw1),
                   'Nonverbal IQ' = mean(MullenRaw1),
                   'Caucasian' = sum(Ethnicity == 'White'),
                   'Non-Caucasian' = sum(Ethnicity != 'White'),
                   'Socialization' = mean(Socialization),
                   'Word used' = mean(tokens_CHI),
                   'unique words' = mean(types_CHI),
                   'MLU Child' = mean(CHI_MLU),
                   'MLU Parent' = mean (MOT_MLU))
```

The participants were categorized in two groups with 29 ASD participants and 32 TD participants. The participants were mostly Caucasian (53 out of 61) where the majority were male (51 out of 61). There was a higher percentage of Non-Caucasian in the ASD group. The mean age of the first visit was 33 months for the ASD group and only 20 months for TD where the difference was justified due to matching the groups up by verbal IQ where the ASD group had a slightly lower verbal IQ (-3) at the time of the first visit. Non verbal IQ was closely matched at the first visit for the two groups. The ASD group had a lower socialization than the TD children and used less words in general even though the two groups had almost same MLU.


## Let's test hypothesis 1: Children with ASD display a language impairment  (Exercise 2)

### Hypothesis: The child's MLU changes: i) over time, ii) according to diagnosis

Let's start with a simple mixed effects linear model

Remember to plot the data first and then to run a statistical test.
- Which variable(s) should be included as fixed factors?
- Which variable(s) should be included as random factors?

```{r ex2, include = FALSE}

# A plot of the average development of each group in mean length of utterances (CHI_MLU) over visits - each participant with their own color, line showing the average development, one plot for each group
ggplot(data = train,
       aes(x = VISIT, y = CHI_MLU, group = Diagnosis, color = SUBJ)) +
  geom_smooth(method = lm) +
  geom_point() + 
  facet_wrap(.~ Diagnosis)

# Shows that there is a more even development in TD than ASD and also that the average MLU is higher for TD, especially at visit 6. The ASD graph has more variance 

# A plot where there is created a line for each participants estimated development. Shows that there is a great deal more variance in the development of ASD children where most of the TD children develop in the same way.
ggplot(train, aes(x = VISIT, y = CHI_MLU, group = SUBJ, color = SUBJ)) +
  geom_smooth(method = 'lm', se = FALSE) +
  facet_wrap(~Diagnosis) +
  theme_minimal()

# Concluding: ASD has more variance thant TD kids 

## Creating models - Assuming that the intercept does not affect the slop, meaning that where the childs ability starts will not affect the way in which they develop over time.

# model with no interactions - the most basic model
m1 <- lmer(data = train, 
     CHI_MLU ~ VISIT + Diagnosis + (1|SUBJ) + (0 + VISIT|SUBJ))

# interactions - the slope of visits will change based on diagnosis
m2 <- lmerTest::lmer(data = train, 
     CHI_MLU ~ VISIT * Diagnosis + (1|SUBJ) + (0 + VISIT|SUBJ))


### Notes taken in class ###

#Using lmer => random effects 
# => the child ID is important 
# => + (1|ID) => everyone has a different starting point => the unikness of the child only is related to the intersept => all kids will be parallel, where all ASD will be parallel and all TD will be paralel 
# That is why it is not enough 
# adding to it, that the child ID all so might affect visit => variaing effect => (0 + Visit|ID)
# adding (1 + visit | ID) is meaning that the two things are related 


#### Adding Verbal IQ 
# tries to explain some of the variance - add it as an interaction - it does not only affect where you start but also how you develop - meaning it affects you the same no matter if you have ASD or not - this is not what we want so we add it directly (Diagnosis * Visit * Verbal IQ) 

#### Want to add both non verbal iq and socialization => this would create a lot of models

### Want us to find the best posible model
```

How would you evaluate whether the model is a good model?
We want to look for a model that explains a lot of the variance (looking at the R2 margenal) but also has a low AIC value. First we will use anova to see if it explaince more variance and than look at the R2 values


```{r ex2 evaluate, include = FALSE}
summary(m1)

summary(m2)

qqnorm(residuals(m2))

anova(m1, m2)
#model 2 is better than model 1 since it has a lower AIC and also explainse significantly more of the variance 

r.squaredGLMM(m2)
# When looking at the marginal R2 the model only predicts around 35% of the variance but when including the random effects in the conditional R2 we see that almost 82% of the variance is explained. This means that the random effects explain a great deal of the variance, and this will not be good if we want to make a generalized model.
```

Not too good, right? Let's check whether a growth curve model is better.
Remember: a growth curve model assesses whether changes in time can be described by linear, or quadratic, or cubic (or... etc.) components.
First build the different models, then compare them to see which one is better.

```{r ex2 growth curve, include = FALSE}

# Chosen to skip this exercis since we have not covered growth curve

```

Exciting right? Let's check whether the model is doing an alright job at fitting the data. Plot the actual CHI_MLU data against the predictions of the model fitted(model). 

```{r}
# saving train in a new data frame befor changing it
df <- train

#adding the predicted values to each trail based on model 2
df$fit <- predict(m2)

# # predicted versus actual, points are actual datapoints, lines are predictions
ggplot(df, aes(x = VISIT, y = CHI_MLU, group = SUBJ, color = Diagnosis)) +
  geom_point(alpha = 0.7, position = position_jitter(w = 0.1, h = 0)) +
  facet_wrap(~SUBJ) +
  theme_minimal() +
  geom_smooth(method = "lm", se = FALSE, size = 0.4, aes(y=fit)) +
  labs(title = "Model prediction on current data",
       x = "Visit",
       y = "Mean length of utterance",
       color = "Diagnosis")



```

Now it's time to report our results.
Remember to report:
- the estimates for each predictor (beta estimate, standard error, p-value)
- A plain word description of the results
- A plot of your model's predictions (and some comments on whether the predictions are sensible)

[REPORT THE RESULTS]
Linguistic development of children MLU is affected by ... [COMPLETE]

## Let's test hypothesis 2: Parents speak equally to children with ASD and TD  (Exercise 3)

### Hypothesis: Parental MLU changes: i) over time, ii) according to diagnosis

```{r ex3, include = FALSE}
# A plot over the average difference in parents MLU depending on diagnosis 
ggplot(data = train,
       aes(x = VISIT, y = MOT_MLU, group = Diagnosis, color = SUBJ)) +
  geom_smooth(method = lm) +
  geom_point() + 
  facet_wrap(.~ Diagnosis)
# Shows that TD parents on average talks more but it shows they have almost the same slope, meaning they develop in the same way 

# Shows an estimated line for each parent
ggplot(train, aes(x = VISIT, y = MOT_MLU, group = SUBJ, color = SUBJ)) +
  geom_smooth(method = 'lm', se = FALSE) +
  facet_wrap(~Diagnosis) +
  theme_minimal()
# Shows that there is fare more variance in the parents speach than the childrens. The variance can also be seen in the TD parents 

# Assuming that the intercept does not affect the slop 

m3 <-lmer(data = train, 
     MOT_MLU ~ Diagnosis + (1|SUBJ) + (0 + VISIT|SUBJ))

m4 <- lmer(data = train, 
     MOT_MLU ~ VISIT + (1|SUBJ) + (0 + VISIT|SUBJ))

m5 <- lmer(data = train, 
     MOT_MLU ~ VISIT + Diagnosis + (1|SUBJ) + (0 + VISIT|SUBJ))
# no interactions

m6 <- lmerTest::lmer(data = train, 
     MOT_MLU ~ VISIT * Diagnosis + (1|SUBJ) + (0 + VISIT|SUBJ))

summary(m3)

summary(m4)

summary(m5)

summary(m6)

anova(m3, m4, m5,m6) 
# the best model is m5

r.squaredGLMM(m5)


ggplot(data = train,
       aes(x = VISIT, y = MOT_MLU, group = SUBJ, color = Diagnosis)) +
  geom_point(alpha = 0.7, position = position_jitter(w = 0.1, h = 0)) +
  facet_wrap(~SUBJ) +
  theme_minimal() +
  geom_smooth(method = "lm", se = FALSE, size = 0.4, aes(y=fit)) +
  labs(title = "Model prediction on current data",
       x = "Visit",
       y = "Mean length of utterance",
       color = "Diagnosis")
```

The R2 margen for adults shows that visit and diagnosis explains about 20 % of the variance where R2 conditional includes the random effects and explaince about 65%. This suggests that time and diagnosis can affect the Parent MLU but this might not be the only thing.

Parent MLU is affected by ... but probably not ...
[REPORT THE RESULTS]

### Adding new variables (Exercise 4)

Your task now is to figure out how to best describe the children linguistic trajectory. The dataset contains a bunch of additional demographic, cognitive and clinical variables (e.g.verbal and non-verbal IQ). Try them out and identify the statistical models that best describes your data (that is, the children's MLU). Describe how you selected the best model and send the code to run the model to Riccardo and Kenneth


```{r ex4, include = FALSE}
m7 <- lmerTest::lmer(data = train, 
     CHI_MLU ~ VISIT * Diagnosis * ExpressiveLangRaw1+ (1|SUBJ) + (0 + VISIT|SUBJ))

anova(m2, m7)

r.squaredGLMM(m7)
# model 7 is a better model than model 2 

m8 <- lmerTest::lmer(data = train, 
     CHI_MLU ~ VISIT * Diagnosis * ExpressiveLangRaw1 * Socialization1 + (1|SUBJ) + (0 + VISIT|SUBJ))

anova(m7, m8)
r.squaredGLMM(m8)

# This is a better model but it might be over fitted 

m9 <- lmerTest::lmer(data = train, 
     CHI_MLU ~ 1 + VISIT * Diagnosis * MOT_MLU * ExpressiveLangRaw1 + (1|SUBJ) + (0 + VISIT|SUBJ))

summary(m9)

anova(m8,m9)
r.squaredGLMM(m9)

m10 <- lmer(data = train, 
     CHI_MLU ~ VISIT * Diagnosis * MOT_MLU * ExpressiveLangRaw1 + (1 + VISIT|SUBJ))

anova(m9,m10)

# m9 Explains the most

m11 <- lmerTest::lmer(data = train, 
     CHI_MLU ~ VISIT * Diagnosis * MOT_MLU + (1|SUBJ) + (0 + VISIT|SUBJ))

anova(m9,m11)

m12 <- lmer(data = train, 
     CHI_MLU ~ 1 + VISIT * Diagnosis + VISIT * MOT_MLU + VISIT * ExpressiveLangRaw1 +  (1|SUBJ) + (0 + VISIT|SUBJ))

r.squaredGLMM(m12) ###### The model we will use
anova(m9,m12)

m13 <- lmerTest::lmer(data = train, 
     CHI_MLU ~ 1 + VISIT * ADOS1 * MOT_MLU * ExpressiveLangRaw1 + (1|SUBJ) + (0 + VISIT|SUBJ))

r.squaredGLMM(m13)

anova(m9, m13)

m14 <- lmerTest::lmer(data = train, 
     CHI_MLU ~ 1 + VISIT * Diagnosis * ExpressiveLangRaw1 + ExpressiveLangRaw1 * MOT_MLU + (1 + VISIT|SUBJ))

r.squaredGLMM(m14)

anova(m9, m14)

######## Notes from class ##############

#3-way interactions are hard to interpret but explains the most - in a case of just wanting to being able to predict the new data 

# Theory 
# Think about what parameters affect on the language development before you look at the data => that is what you are going to test => publiching it before running the experiment => does not allow you to explore the data 

# Lasso
# using a max model => putting in all the variables : Taking all the data and removing all the beta that is between -1 and 1 and exclude those values and runs it again until there no longer is any that fall between -1 and 1 
# dividing all the betas with the sum of the betas 
# a small effect is more likely to be overfitting 
# letting he data talk for its own

# Doing something in between 
# finding some good variables and models

```

In addition to ..., the MLU of the children is also correlated with ...
Using AIC / nested F-tests as a criterium, we compared models of increasing complexity and found that ...


Our best model was model 7 but hard to explain 
Otherwise model 10 was a great alternative that is easier to explain

[REPORT THE RESULTS]

```{r}
# Applying the function from part 2 to the original data 
demo_train <- read.csv('demo_train.csv', header = T, sep =',')
token_train <- read.csv('token_train.csv', header = T, sep =',')
LU_train <- read.csv('LU_train.csv', header = T, sep = ',')
```

