---
title: "Assignment 2 - Language Development in ASD - Making predictions"
author: "Riccardo Fusaroli"
date: "August 9, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome to the second exciting part of the Language Development in ASD exercise

In this exercise we will delve more in depth with different practices of model comparison and model selection, by first evaluating your models from last time against some new data. Does the model generalize well?
Then we will learn to do better by cross-validating models and systematically compare them.

The questions to be answered (in a separate document) are:
1- Discuss the differences in performance of your model in training and testing data
2- Which individual differences should be included in a model that maximizes your ability to explain/predict new data?
3- Predict a new kid's performance (Bernie) and discuss it against expected performance of the two groups

## Learning objectives

- Critically appraise the predictive framework (contrasted to the explanatory framework)
- Learn the basics of machine learning workflows: training/testing, cross-validation, feature selections

## Let's go

N.B. There are several datasets for this exercise, so pay attention to which one you are using!

1. The (training) dataset from last time (the awesome one you produced :-) ).
2. The (test) datasets on which you can test the models from last time:
* Demographic and clinical data: https://www.dropbox.com/s/ra99bdvm6fzay3g/demo_test.csv?dl=1
* Utterance Length data: https://www.dropbox.com/s/uxtqqzl18nwxowq/LU_test.csv?dl=1
* Word data: https://www.dropbox.com/s/1ces4hv8kh0stov/token_test.csv?dl=1

### Exercise 1) Testing model performance

How did your models from last time perform? In this exercise you have to compare the results on the training data () and on the test data. Report both of them. Compare them. Discuss why they are different.

- recreate the models you chose last time (just write the model code again and apply it to your training data (from the first assignment))
- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the function rmse())
- create the test dataset (apply the code from assignment 1 to clean up the 3 test datasets)
- test the performance of the models on the test data (Tips: google the functions "predict()")
- optional: predictions are never certain, can you identify the uncertainty of the predictions? (e.g. google predictinterval())


```{r, include = FALSE}

pacman::p_load(readr,dplyr,stringr,lmerTest,Metrics,caret, MuMIn)
pacman::p_load(rlist)

#Training set
demo_train <- read.csv('demo_train.csv', header = T, sep =',')
token_train <- read.csv('token_train.csv', header = T, sep =',')
LU_train <- read.csv('LU_train.csv', header = T, sep = ',')

#Testing set
demo_test <- read.csv('demo_test.csv', header = T, sep =',')
token_test <- read.csv('token_test.csv', header = T, sep =',')
LU_test <- read.csv('LU_test.csv', header = T, sep = ',')

## Clean up function, included to inspire you

CleanUpData <- function(Demo,LU,Word){
  
  Speech <- merge(LU, Word) %>% 
    rename(
      Child.ID = SUBJ, 
      Visit=VISIT) %>%
    mutate(
      Visit = as.numeric(str_extract(Visit, "\\d")),
      Child.ID = gsub("\\.","", Child.ID)
      ) %>%
    dplyr::select(
      Child.ID, Visit, MOT_MLU, CHI_MLU, types_MOT, types_CHI, tokens_MOT, tokens_CHI
    )
  
  Demo <- Demo %>%
    dplyr::select(
      Child.ID, Visit, Ethnicity, Diagnosis, Gender, Age, ADOS, MullenRaw, ExpressiveLangRaw, Socialization
    ) %>%
    mutate(
      Child.ID = gsub("\\.","", Child.ID)
    )
    
  Data=merge(Demo,Speech,all=T)
  
  Data1= Data %>% 
     subset(Visit=="1") %>% 
     dplyr::select(Child.ID, ADOS, ExpressiveLangRaw, MullenRaw, Socialization) %>%
     rename(Ados1 = ADOS, 
            verbalIQ1 = ExpressiveLangRaw, 
            nonVerbalIQ1 = MullenRaw,
            Socialization1 = Socialization) 
  
  Data=merge(Data, Data1, all=T) %>%
    mutate(
      Child.ID = as.numeric(as.factor(as.character(Child.ID))),
      Visit = as.numeric(as.character(Visit)),
      Gender = recode(Gender, 
         "1" = "M",
         "2" = "F"),
      Diagnosis = recode(Diagnosis,
         "A"  = "TD",
         "B"  = "ASD")
    )

  return(Data)
}

# Load training Data

#- recreate the models you chose last time (just write the code again and apply it to Train Data)

#- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the function rmse())

#- create the test dataset (apply the code from assignment 1 or my function to clean up the 3 test datasets)
# Test data

#- test the performance of the models on the test data (Tips: google the functions "predict()")

#- optional: predictions are never certain, can you identify the uncertainty of the predictions? (e.g. google predictinterval())

train <- CleanUpData(demo_train, LU_train, token_train)

test <- CleanUpData(demo_test,LU_test,token_test)

train <- subset(train, !is.na(CHI_MLU))

test <- subset(test, !is.na(CHI_MLU))

test$Child.ID <- test$Child.ID + 66


# simplest model - C_MLU_model
C_MLU_model <- lmer(data = train, 
     CHI_MLU ~ 1 + Visit * Diagnosis + Visit * MOT_MLU + Visit * verbalIQ1 + (1|Child.ID) + (0 + Visit|Child.ID))

predict_model <- predict(C_MLU_model)

#- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the function rmse())
rmse(train$CHI_MLU, predict_model) # RMSE = 0.35

#- test the performance of the models on the test data (Tips: google the functions "predict()")
predict_model_test <- predict(C_MLU_model, test, allow.new.levels = TRUE)
rmse(test$CHI_MLU, predict_model_test) # RMSE = 0.43 


## MODEL 10
m10 <- lmer(data = train, 
     CHI_MLU ~ Visit * Diagnosis * MOT_MLU * verbalIQ1 +(1|Child.ID) + (0 + Visit|Child.ID))

predict_m10 <- predict(m10)

#- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the function rmse())
rmse(train$CHI_MLU, predict_m10) # RMSE = 0.34

#- test the performance of the models on the test data (Tips: google the functions "predict()")
predict_m10_test <- predict(m10, test, allow.new.levels = TRUE)
rmse(test$CHI_MLU, predict_m10_test) # RMSE = 0.40

```

So, the model prediction performance on the test data for our prefered model (C_MLU_model) shows a root mean squared error of 0.35 MLU which is higher than the error on the train data (0.43).

Therefore, we might have over fitted the model on the train data and it is not as good at predicting on novel data. On the other hand it can still be argued that it is an acceptal model depending on how conservative you are. 

For model 10 we see that there is almost no difference from the first model (0.34 MLU for the training set and 0.40 MLU for the testing set). This model is on the other hand easier to interpret.


### Exercise 2) Model Selection via Cross-validation (N.B: ChildMLU!)

One way to reduce bad surprises when testing a model on new data is to train the model via cross-validation. 

In this exercise you have to use cross-validation to calculate the predictive error of your models and use this predictive error to select the best possible model.

- Use cross-validation to compare your model from last week with the basic model (Child MLU as a function of Time and Diagnosis, and don't forget the random effects!)
- (Tips): google the function "createFolds";  loop through each fold, train both models on the other folds and test them on the fold)


- Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.

- Bonus Question 1: What is the effect of changing the number of folds? Can you plot RMSE as a function of number of folds?
- Bonus Question 2: compare the cross-validated predictive error against the actual predictive error on the test data


```{r}
### Combining the test and training set ### 
#train <- rbind(train, test)

#- Create the basic model of ChildMLU as a function of Time and Diagnosis (don't forget the random effects!).
m0 <- lmer(data = train, CHI_MLU ~ 1 + Visit + Diagnosis + (1|Child.ID) + (0 + Visit | Child.ID))

#- Make a cross-validated version of the model. (Tips: google the function "createFolds";  loop through each fold, train a model on the other folds and test it on the fold)

###### Creating our own loop #########

# loop assigning folds to dataframe by participant and diagnosis
for (id in unique(train$Child.ID)){ # loop through each unique participant
  
  ### Run at first iteration (setup variables)
  if (id == min(unique(train$Child.ID))) {
    # create folds for each diagnosis
    # list = FALSE returns vector instead of list
    fold_list_TD <- createFolds(unique(train$Child.ID[train$Diagnosis == "TD"]), k = 5, list = FALSE)
    fold_list_ASD <- createFolds(unique(train$Child.ID[train$Diagnosis == "ASD"]), k = 5, list = FALSE)
    
    # identifiers for counting values in folds
    fold_i_TD <- 1
    fold_i_ASD <- 1 }
  
  
  ### Run at every iteration (assign folds to dataframe)
  if (unique(train$Diagnosis[train$Child.ID == id]) == "TD") { # if participant has diagnosis == TD
    train$fold[train$Child.ID == id] <- fold_list_TD[fold_i_TD] # assign fold-value from fold_list_TD vector
    # update identifier
    fold_i_TD <- fold_i_TD + 1 }
  else {
    train$fold[train$Child.ID == id] <- fold_list_ASD[fold_i_ASD]
    fold_i_ASD <- fold_i_ASD + 1 }
  
  
  ### Run at last iteration (print information and cleanup)
  if (id == max(unique(train$Child.ID))) {
    message(paste(""))
    # this is just for checking distribution, can be deleted, also why it's not commented
    for (i in 1:max(train$fold)) {
      TD_in_fold <- length(unique(train$Child.ID[train$fold == i & train$Diagnosis == "TD"]))
      ASD_in_fold <- length(unique(train$Child.ID[train$fold == i & train$Diagnosis == "ASD"]))
      percent_TD <- ceiling((TD_in_fold / (TD_in_fold + ASD_in_fold)) * 100)
      percent_ASD <- floor((ASD_in_fold / (TD_in_fold + ASD_in_fold)) * 100)
      message(paste("n for fold:", i, "  TD:", TD_in_fold, "  ASD:", ASD_in_fold, "  %TD / %ASD :", percent_TD, "/", percent_ASD))
      if (i == max(train$fold)) {
        total_TD <- length(unique(train$Child.ID[train$Diagnosis == "TD"]))
        total_ASD <- length(unique(train$Child.ID[train$Diagnosis == "ASD"]))
        percent_TD <- (total_TD / (total_TD + total_ASD)) * 100
        percent_ASD <- (total_ASD / (total_TD + total_ASD)) * 100
        message(paste(""))
        message(paste("Total %TD / %ASD :", percent_TD, "/", percent_ASD))
        rm(TD_in_fold, ASD_in_fold, percent_TD, percent_ASD, total_ASD, total_TD, i) }
    }
    # cleanup enviroment
    rm(fold_i_ASD, fold_i_TD, fold_list_ASD, fold_list_TD, id) }
}


# loop through each fold, training on k-1 folds and testing on the remaining fold
for (i in 1:max(train$fold)) { # run loop k times
  
  # initialize performance vector for registering RMSE performances if i == 1
  if (i == 1){
    performance <- c() }
  
  # init empty prediction vector for each iteration
  prediction <- c()
  # initialize empty train and testing sets for each iteration
  train_fold <- as.data.frame(train[0,])
  test_fold <- as.data.frame(train[0,])
  fold_iteration <- as.integer(as.character(i)) # convert i to integer
  
  # assign test and training folds depending on current identifier value
  train_fold <- train[train$fold != fold_iteration,]
  test_fold <- train[train$fold == fold_iteration,]
  
  # model setup
  m1 <- lmer(data = train_fold, CHI_MLU ~ 1 + Visit + Diagnosis + (1|Child.ID) + (0 + Visit | Child.ID))
  
  # run prediction
  prediction <- predict(m1, test_fold, allow.new.levels = TRUE) # run prediction on testing fold
  performance[i] <- rmse(test_fold$CHI_MLU, prediction) # calculate rmse for each fold and append to performance vector
  #fold_rmse <- rmse(test_fold$CHI_MLU, prediction) # calculate rmse for each fold
  #performance[i] <- fold_rmse # register rmse value to 'performances'vector
  message(paste("RMSE for fold", fold_iteration, performance[i])) # print RMSE for each fold to console
  
  if (i == max(train$fold)) { # if it is the last iteration, calculate, print and save mean RMSE
    performance = mean(performance)
    message(paste(""))
    message(paste("Mean RMSE for model:", performance)) # print mean RMSE score
    # remove temporary variables
    rm(test_fold, train_fold, fold_iteration, prediction, m1) }
}




###### WITH THE CVMS PACKAGE ######

pacman::p_load(cvms, groupdata2, knitr, dplyr, ggplot2)
newdata = train
data <- fold(newdata, k = 5,
             cat_col = 'Diagnosis',
             id_col = 'Child.ID') %>% 
  arrange(.folds)

data$Diagnosis <- as.factor(data$Diagnosis)


cross_m0 <- cross_validate(data, "CHI_MLU ~ 1 + Visit + Diagnosis + (1|Child.ID) + (0 + Visit|Child.ID)", 
               fold_cols =".folds",
               family = "gaussian", 
               REML = FALSE)

cross_m0$model <- c('Basic model')



cross_m1 <- cross_validate(data, "CHI_MLU ~ 1 + Visit * Diagnosis + Visit * MOT_MLU + Visit * verbalIQ1 + (1|Child.ID) + (0 + Visit|Child.ID)", 
               fold_cols =".folds",
               family = "gaussian", 
               REML = FALSE)

cross_m1$model <- c('model 2')

cross_m2 <- cross_validate(data, "CHI_MLU ~ 1 + Visit * Diagnosis  * MOT_MLU  * verbalIQ1 + (1|Child.ID) + (0 + Visit|Child.ID)", 
               fold_cols =".folds",
               family = "gaussian", 
               REML = FALSE)

cross_m2$model <- c('model 3')

Model_overview <- rbind(cross_m0,cross_m1)
Model_overview <- rbind(Model_overview, cross_m2)
```

Above, we have compared our C_MLU_model with the basic model and found that it has a RMSE of 0.57 and the basic model has a higer RMSE of 0.86. That means that our model is a better predicter than the basic model.
On the other hand we have also conducted cross validation of the more complecated model (m10) and found that it has a lower RMSE (0.54) and a lower AIC than the C_MLU_model. The problem is still that it is a more complecated model that is hard to interpret so it is a question if it is worth the traidoff 


### Exercise 3) Assessing the single child

Let's get to business. This new kiddo - Bernie - has entered your clinic. This child has to be assessed according to his group's average and his expected development.

Bernie is one of the six kids in the test dataset, so make sure to extract that child alone for the following analysis.

You want to evaluate:

- how does the child fare in ChildMLU compared to the average TD child at each visit? Define the distance in terms of absolute difference between this Child and the average TD.

- how does the child fare compared to the model predictions at Visit 6? Is the child below or above expectations? (tip: use the predict() function on Bernie's data only and compare the prediction with the actual performance of the child)

```{r}
### Bernie is Child.ID 68 #### compared the original dataset with the anominized
train <- rbind(train, test)

Bernie <- train[train$Child.ID == 68,]

Total_data <- train[train$Child.ID != 68,]

Total_data_TD <- Total_data[Total_data$Diagnosis == 'TD',]

# how does the child fare in ChildMLU compared to the average TD child at each visit? 

for (i in 1:6) {
  visit <- Total_data_TD[Total_data_TD$Visit == i,]
  bernie_difference <- mean(visit$CHI_MLU) - mean(Bernie$CHI_MLU[Bernie$Visit == i])
  message(paste("Difference in CHI_MLU for Bernie at Visit", i, ": ", bernie_difference , ':', mean(visit$CHI_MLU)))
  
  if (i == 6){
    rm(visit, bernie_difference) }
}

# Define the distance in terms of absolute difference between this Child and the average TD.
mean(Total_data_TD$CHI_MLU) - mean(Bernie$CHI_MLU)


#how does the child fare compared to the model predictions at Visit 6?
Bpredictions <- predict(C_MLU_model, Bernie, allow.new.levels = TRUE)
rmse(Bernie$CHI_MLU[Bernie$Visit == 6], Bpredictions)

Bpredictions[6] - Bernie$CHI_MLU[6]



## Comparing to ASD

Total_data_ASD <- Total_data[Total_data$Diagnosis == 'ASD',]

for (i in 1:6) {
  visit <- Total_data_ASD[Total_data_ASD$Visit == i,]
  bernie_difference <- mean(visit$CHI_MLU) - mean(Bernie$CHI_MLU[Bernie$Visit == i])
  message(paste("Difference in CHI_MLU for Bernie at Visit", i, ": ", bernie_difference , ':', mean(visit$CHI_MLU)))
  
  if (i == 6){
    rm(visit, bernie_difference) }
}
```


[HERE GOES YOUR ANSWER]

### OPTIONAL: Exercise 4) Model Selection via Information Criteria
Another way to reduce the bad surprises when testing a model on new data is to pay close attention to the relative information criteria between the models you are comparing. Let's learn how to do that!

Re-create a selection of possible models explaining ChildMLU (the ones you tested for exercise 2, but now trained on the full dataset and not cross-validated).

Then try to find the best possible predictive model of ChildMLU, that is, the one that produces the lowest information criterion.

- Bonus question for the optional exercise: are information criteria correlated with cross-validated RMSE? That is, if you take AIC for Model 1, Model 2 and Model 3, do they co-vary with their cross-validated RMSE?

### OPTIONAL: Exercise 5): Using Lasso for model selection
Welcome to the last secret exercise. If you have already solved the previous exercises, and still there's not enough for you, you can expand your expertise by learning about penalizations. Check out this tutorial: http://machinelearningmastery.com/penalized-regression-in-r/ and make sure to google what penalization is, with a focus on L1 and L2-norms. Then try them on your data!

